---
title: "How to use the wqsPerm package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction 

*(Taken from Drew's paper)*

Weighted quantile sum regression is a statistical technique to evaluate the effect of complex exposure mixtures on an outcome (Carrico et al. 2015). It is a single-index method which estimates a combined mixture sum effect as well as weights determining each individual mixture component's contributions to the sum effect. However, the model features a statistical power and Type I error (i.e., false positive) rate tradeoff, as there is a machine learning step to determine the weights that optimize the linear model fit. If the full data is used to estimate both the mixture component weights and the regression coefficients, there is high power but also a high false positive rate since coefficient p-values are calculated for a weighted mixture independent variable calculated using weights that have already been optimized to find a large effect. 

We recently proposed an alternative method based on a permutation test that should reliably allow for both high power and low false positive rate when utilizing the WQSr. The permutation test is a method of obtaining a p-value by simulating the null distribution through permutations of the data. The permutation test algorithm is described more in detail in (reference Drew's paper).

**[Needs a brief technical intro to the permutation test.]**

## How to use the `wqsPerm` package 

The `wqsPerm` package builds from the `gWQS` package.

The two main functions of the `wqsPerm` package are `wqsperm` and `wqsfullperm`. 

### `wqsperm`

`wqsperm` uses a `gwqs` object (from the `gWQS` package) as an input. To use `wqsperm`, we first need to create a `gwqs` object for our *reference WQS* run, setting `validation = 0`, designating the number of bootstraps,  and choosing a random seed to use for the remaining WQS function calls in the algorithm. 

Next, we need to create a second `gwqs` object for our *permutation test WQS* run with the same random seed and again setting `validation = 0`. However, for this second `gwqs` object, we may elect a bootstrap count lower than that in the reference object for the sake of efficiency. 

We will use this second `gwqs` object as an input for the `wqsperm` package. We will put this object in the `model` `gwqs` argument and set the following parameters: 

* `niter`: Number of permutation test iterations 
* `boots`: Number of bootstraps for the WQS run in each permutation test iteration 
* `b1_pos`: Whether beta_1 is restricted to the positive direction 
* `rs`: Random subset implementation 

**Important Note**: The arguments `b1_pos` and `rs` should be consistent with the inputs chosen in the `model` object.   

The permutation test will output the following in a list: 

* `pval`: p-value for the proportion of permuted WQS coefficient values greater than the reference value.  
* `testbeta`: reference beta1 value 
* `betas`: a vector of beta values from each iteration of the permutation test 

### `wqsfullperm` 

The second function `wqsfullperm` is a full wrapper function which implements the steps from the previous section. 

To use `wqsfullperm`, you must specify the same required arguments as needed in the `gwqs` call. For the bootstrap count `b` argument, you must specify `b_main`, the number of bootstraps for the **reference WQS** run and `b_perm`, the number of bootstraps for the **permutation test WQS** run. As with before, it is acceptable/standard for `b_main` $>$ `b_perm` for the sake of efficiency. Finally, you should indicate the number of desired permutation test runs `niter`. 

### Limitations 

Note that there are limitations in the current version of the WQS permutation test algorithm and package implementation: 

* **Linear regression only:** The permutation test algorithm can only `gwqs` inputs where `family = gaussian`. A logistic regression version for the algorithm is currently in the works. 
* **Interaction terms or stratified weights:**  This permutation test is not yet set up to accomodate stratified weights or WQS interaction terms.

## Example 1 (using `wqsperm`)

```{r, warning = F, message = F}
library(gWQS)
library(wqsperm)

# mixture names 
PCBs <- names(wqs_data)[1:34]

# create reference wqs object with 10 bootstraps 
wqs_ref <- gwqs(yLBX ~ wqs, mix_name = PCBs, data = wqs_data, q = 10, validation = 0, 
                b = 10, b1_pos = T, bl_constr = F, family = "gaussian", seed = 16)

# create permutation test wqs object with 5 bootstraps
wqs_perm <- gwqs(yLBX ~ wqs, mix_name = PCBs, data = wqs_data, q = 10, validation = 0, 
                b = 5, b1_pos = T, bl_constr = F, family = "gaussian", seed = 16)

# run permutation test
perm_test_res <- wqsperm(wqs_perm, niter = 10, boots = 5, b1_pos = T)

# report reference wqs results, with permutation test p-value 
perm_test_res$pval
```

## Example 2 (using `wqsfullperm`)

```{r}
perm_test_res <- wqsfullperm(formula = yLBX ~ wqs, 
                             data = wqs_data, 
                             mix_name = PCBs, 
                             q = 10, 
                             b_main = 10, 
                             b_perm = 5, 
                             b1_pos = T, 
                             niter = 10, 
                             seed = 16, 
                             plan_strategy = "multicore", 
                             returnbetas = TRUE)
```
