---
title: "How to use the wqsPerm package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction 

*(Taken from Drew's paper)*

Weighted quantile sum regression is a statistical technique to evaluate the effect of complex exposure mixtures on an outcome (Carrico et al. 2015). It is a single-index method which estimates a combined mixture sum effect as well as weights determining each individual mixture component's contributions to the sum effect. However, the model features a statistical power and Type I error (i.e., false positive) rate tradeoff, as there is a machine learning step to determine the weights that optimize the linear model fit. If the full data is used to estimate both the mixture component weights and the regression coefficients, there is high power but also a high false positive rate since coefficient p-values are calculated for a weighted mixture independent variable calculated using weights that have already been optimized to find a large effect. 

We recently proposed an alternative method based on a permutation test that should reliably allow for both high power and low false positive rate when utilizing WQS. The permutation test is a method of obtaining a p-value by simulating the null distribution through permutations of the data. The permutation test algorithm is described more in detail in (reference Drew's paper).

## How to use the `wqsperm` package 

The `wqsperm` package builds from the `gWQS` package.

The two main functions of the `wqsPerm` package are `wqs_perm` and `wqs_full_perm`. 

### `wqs_perm`

`wqs_perm` uses a `gwqs` object (from the `gWQS` package) as an input. To use `wqs_perm`, we first need to run an initial *WQS regression* while setting `validation=0`.  

We will use this `gwqs` object as the `model` argument for the `wqs_perm` function and set the following additional parameters: 

* `boots`: Number of bootstraps for the WQS run in each permutation test iteration. Note that we may elect a bootstrap count `boots` lower than that specified in the `model` object for the sake of efficiency. In this case, `wqs_perm` will generate a permutation test reference WQS run with that number of bootstraps to determine the reference beta_1 value. 
* `niter`: Number of permutation test iterations 
* `b1_pos`: Whether beta_1 is restricted to the positive direction 
* `rs`: Random subset implementation 
* `plan_strategy`: Evaluation strategy for the plan function 
* `seed`: Random seed 

The arguments `b1_pos` and `rs` should be consistent with the inputs chosen in the `model` object. The `seed` should ideally be consistent with the seed set in the `model` object, though this is not required. 

The permutation test will output the following in a nested list: 

* `perm_test`
  * `pval`: p-value for the proportion of permuted WQS coefficient values greater than the reference value.  
  * `testbeta`: reference beta1 value 
  * `betas`: a vector of beta values from each iteration of the permutation test
* `gwqs_main`: main gWQS object (same as `model` input)
* `gwqs_perm`: permutation test reference gWQS object (NULL if same number of bootstraps)

### `wqs_full_perm` 

The second function `wqs_full_perm` is a full wrapper which implements the initial gWQS run and permutation test in one function call. 

To use `wqs_full_perm`, you must specify the same required arguments as needed in the `gwqs` call. For the bootstrap count `b` argument, you must specify `b_main`, the number of bootstraps for the *reference WQS regression* run and `b_perm`, the number of bootstraps for the *permutation test WQS regression* run. As with before, it is acceptable/standard for `b_main` $>$ `b_perm` for the sake of efficiency. Finally, you should indicate the number of desired permutation test runs `niter`.

The outputs of the `wqs_full_perm` function are the same as those in the `wqs_perm` function. 

### Limitations 

Note that there are limitations in the current version of the WQS permutation test algorithm and package implementation: 

* **Linear regression only:** The permutation test algorithm can only `gwqs` inputs where `family = gaussian` and `link = identity`. A logistic regression version for the algorithm is currently in the works. 
* **Interaction terms or stratified weights:**  This permutation test is not yet set up to accomodate stratified weights or WQS interaction terms.

## Example 1 (using `wqs_perm`)

```{r, warning = F, message = F}
library(gWQS)
library(wqsperm)

# mixture names
PCBs <- names(wqs_data)[1:34]

# TODO: Change bootstrap/iteration counts to be more realistic

# create reference wqs object with 1000 bootstraps
wqs_ref <- gwqs(yLBX ~ wqs, mix_name = PCBs, data = wqs_data, q = 10, validation = 0,
                b = 10, b1_pos = T, plan_strategy = "multicore", family = "gaussian", seed = 16)

# run permutation test
perm_test_res <- wqs_perm(wqs_ref, niter = 10, boots = 5, b1_pos = T)
```

```{r}
print(perm_test_res)
```


## Example 2 (using `wqsfullperm`)

```{r}
# perm_test_res <- wqsfullperm(formula = yLBX ~ wqs, 
#                              data = wqs_data, 
#                              mix_name = PCBs, 
#                              q = 10, 
#                              b_main = 10, 
#                              b_perm = 5, 
#                              b1_pos = T, 
#                              niter = 10, 
#                              seed = 16, 
#                              plan_strategy = "multicore", 
#                              returnbetas = TRUE)
# 
# perm_test_res
```
